---
title: "Class 09:Mini Project"
author: "Anyoleth Alarcon PID: A17347293"
format: pdf
toc: true
---

Today, we will take a step back to some data we can taste, and explore the correlation structure and principal components of some Halloween candy.
## 1. Data Import

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
  
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

Q3. What is your favorite candy in the dataset and what is it's `winpercent` value?

```{r}

# rownames(candy)
# I did `rownames` to see which candies are in the dataset

candy["Hershey's Milk Chocolate",]$winpercent
```

> Q4. What is the `winpercent` value for "Kit Kat"?

```{r}
candy["Kit Kat",]$winpercent
```

Q5. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

## 2. Explanatory Analysis

Want a package to give a quick overview of given dataset. Let's install the `skimr` package and use it on the `candy` data. This can be useful for the first time you encounter a new data set.

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

For `candy$winpercent`, it seems to be on a different scale as it surpasses the scale from 0 to 1.

```{r}
apply(candy, 2, sd)
```


> Q7. What do you think a zero and one represent for the `candy$chocolate` column?

In the *chocolate* column, a zero represents that the candy is not chocolate; and the one represents that the candy is chocolate. AKA 1 = TRUE and 0 = FALSE.

Now we want to start with exploratory analysis w/ histogram.

> Q8. Plot a histogram of `winpercent` values.

```{r}
hist(candy$winpercent, col= "lightblue")
```

> Q9. Is the distribution of `winpercent` values symmetrical?

No.

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

As seen, median is below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.inds <- candy$chocolate ==1
choc.candy <- candy[choc.inds,]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.inds <- candy$fruity ==1
fruit.candy <- candy[fruit.inds,]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```

> Q12. Is this difference statistically significant?

```{r}
ans <- t.test(choc.win, fruit.win)
```

Yes, with a p-value of 2.8713778 x 10^-8 meaning the differences in average win percents are statistically significant.

> Q13. What are the five least liked candy types in this set?

There are 2 related functions that can help here. One is the classic `sort()` and `order()` 

```{r}
x<- c(5,10,1,4)
sort(x)
```

```{r}
order(x)
```

```{r}
inds <- order(candy$winpercent)
head(candy[inds,], 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[inds,],5)
```

```{r}
inds <- order(candy$winpercent, decreasing=T)
head(candy[inds,], 5)
```

> Q15. Make a first barplot of candy ranking based on `winpercent` values.

Make a bar plot with ggplot and order it by winpercent values.

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent))+
  geom_col()
```

Let's add some color now.

```{r}
library(ggplot2)
ggplot(candy) +
  aes(x = winpercent, y = reorder(rownames(candy), winpercent), fill = chocolate) +
  geom_col()
```

Here we want a custom color vector to color each bar the way we want - with chocolate and fruity candy together with whether it is a bar or not.

```{r}
mycols <- rep("gray", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$fruity)] <- "pink"
mycols[as.logical(candy$bar)] <- "brown"

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)
```

```{r}
ggsave("mybarplot.png", width = 3, height = 10)
```

To inset an image with a preffered size:

> Q17. What is the worst ranked chocolate candy?

Charleston Chew

> Q18. What is the best ranked fruity candy?

Starburst

## 3. Winpercent vs Pricepercent

```{r}
# Pink and gray are too light, lets change to red and black
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$fruity)] <- "red"
mycols[as.logical(candy$bar)] <- "brown"
library(ggrepel)
# How about a plot of price vs win
ggplot(candy) +
aes(winpercent, pricepercent, label=rownames(candy)) +
geom_point(col=mycols) +
geom_text_repel(col=mycols, size=3.3, max.overlaps = 9)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for you buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Top 5 most expensive types:

```{r}
inds <- order(candy$price)
tail(candy[inds,], 5)
```

Least popular candy:

```{r}
inds <- order(candy$winpercent)
head(candy[inds,], 1)
```

## 4. Correlation Structure

```{r}
cij <- cor(candy)
cij
```

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are negatively correlated

```{r}
round(cij["chocolate", "winpercent"], 2)
```

## 5. Principal Component Analysis (PCA)

We need to be sure to scale our input candy data before PCA as we have the `winpercent` column on a different scale to all others in the data set.

```{r}
pca <- prcomp(candy, scale. = TRUE)
summary(pca)
```

First main result figure is my "PCA plot"

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x))+
  geom_point(col=mycols)+
  geom_text_repel(max.overlaps=6, col=mycols)+
  theme_bw()
```
The second main PCA result is in the `pca$rotation` we can plot this to generate a so-called "loadings" plot.

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```

```{r}
ggplot(pca$rotation)+
  aes(PC1, reorder(rownames(pca$rotation), PC1), fill = PC1)+
  geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Pluribus, hard, and fruity.
