---
title: "Class 11: AlphaFold"
author: "Anyoleth Alarcon A17347293"
format: pdf
toc: true
---

Here we read the results from AlphaFOld and try to interpret all the models and quality score metrics:

```{r}
library(bio3d)
pth <- "dimer_23119"
pdb.files <- list.files(path = pth, full.names = TRUE, pattern = ".pdb")
```

Align and supperpose all these models

```{r}
file.exists(pdb.files)
```

```{r}
pdbs <- pdbaln(pdb.files, fit = TRUE, exefile="msa")
```

```{r}
plot(pdbs$b[1,], typ="l", ylim=c(0,100), ylab="pLDDT score")
lines(pdbs$b[2,], typ="l", col="blue")
lines(pdbs$b[3,], typ="l", col="green")
lines(pdbs$b[4,], typ="l", col="orange")
lines(pdbs$b[5,], typ="l", col="red")
```

```{r}
rd <- rmsd(pdbs, fit=T)

library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

## Predicted Alignment Error for Domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=pth,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)

pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae2 <- read_json(pae_files[2],simplifyVector = TRUE)
pae3 <- read_json(pae_files[3],simplifyVector = TRUE)
pae4 <- read_json(pae_files[4],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
head(pae1$plddt)
```

```{r}
pae1$max_pae
```
```{r}
pae3$max_pae
```

```{r}
pae5$max_pae
```

We can plot the N by N PAE scores 

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

# Score Residue Conservation from alignment file

AlphaFold returns it's large alignment file used for analysis
Here we read this file and score conservation per position

```{r}
aln_file <- list.files(path=pth,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

```{r}
dim(aln$ali)
```

```{r}
sim <- conserv(aln)
```

```{r}
plotb3(sim[1:99],
       ylab="Conservation Score")
```

Find the consensus sequence at a very high cut-off to find invarient values

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```





